Secure Composition \\ of Robust and Optimising Compilers \vspace{-1.5em}

### 2024-05-28 csf2/main.tex
7

This paper uses a general notion of robust preservation~\cite{abate2021extacc} that considers compilers that use languages with different trace models. 
To this end, considering a source trace $\src{\trace}$ and a target trace $\trg{\trace}$, there is a relation ($\sim$) describing the effect of a corresponding compiler (see \Cref{sec:casestud:rtp}). 
This relation induces the following two projection functions: (1) the \emph{existential image} $\tau_\sim\left(\src{\pi}\right)$ and (2) the \emph{universal image} $\sigma_\sim\left(\trg{\pi}\right)$.
These projections map source-level (resp. target-level) properties to target-level (resp. source-level) properties in a way that identifies the ``same'' property across languages. 
The case study of this paper exclusively uses the universal image, since some considered properties, such as \gls*{ss}, are not definable in a higher-level language that, e.g., does not model speculation.
For self-containment, the universal image is defined as follows~\cite{abate2021extacc}:

\begin{definition}[Universal Image]\label{def:universal:img}\label{def:sigma}
  \[ 
    \sigma_\sim\left(\trg{\pi}\right) := \left\{ \src{\trace} \mid \forall \trg{\trace}, \text{if }\src{\trace}\sim\trg{\trace}, \text{ then } \trg{\trace}\in\trg{\pi} \right\}
  \]
\end{definition}

@@

\subsubsection{\gls*{ms}, Strict Cryptographic Constant Time}

The combination of \gls*{ms} and \gls*{scct} is the intersection of these properties, \gls*{msscct}.
However, \gls*{ms} uses a different trace model than \gls*{scct}, so intersecting them would trivially yield the empty set. 
To remedy this issue, we introduce $\sim_{\ctsafe}: \preevent[_{\ctsafe}] \times \preevent[_{\mssafe}] $, a cross-language trace relation (whose key cases are presented below), that we use to intuitively unify the trace model in which the two properties are expressed:
\[
  \typerulenolabel{mscctrel:drop:tag}{}{\preevent[_{\mssafe}]\sim_{\ctsafe}\preevent[_{\mssafe}];\securitytag{}}
  \typerulenolabel{mscctrel:drop:crash}{}{\lightning\sim_{\ctsafe}\lightning}
\]
\[
  \typerulenolabel{mscctrel:drop:branch}{}{\emptyevent\sim_{\ctsafe}\ev{Branch n};\securitytag{}}
  \typerulenolabel{mscctrel:drop:binop}{}{\emptyevent\sim_{\ctsafe}\ev{Binop n};\securitytag{}}
\]

Essentially, the only work $\sim_{\ctsafe}$ has to do, is to ignore both the new $\ev{Branch\ n}$ and $\ev{Binop\ n}$ base-events as well as the security tag, since they are irrelevant to \gls*{ms}.
Put differently, traces of \gls*{ms} trivially satisfy \gls*{scct}.

@@

\subsubsection{Speculation Memory Safety}

@@

{
\[
  \typerulenolabel{xrel:sms:read}{\trg{n}\text{ in bounds}}{\trg{Get\ \loc\ n;\comp}\sim^{\Ltrg}_{\Lms}\irl{Get\ \loc\ n;\comp}}
\]
\[
  \typerulenolabel{xrel:sms:panic}{\trg{n}\text{ out of bounds}}{\trg{Get\ \loc\ n;\comp}\sim^{\Ltrg}_{\Lms}\irl{\lightning}}
\]
}

@@

This section provides additional insights for the formal development of the paper.
First, this section discusses how to connect each language-specific security property to the general security properties presented in \Cref{sec:compprop} (\Cref{subsec:formalities:maps}), then it demonstrates that the security property resulting from the universal image projection is faithful to the original property (\Cref{subsec:formalities:props}). 
Finally, this section discusses why the order of compiler passes matters, and how does our framework help with identifying insecure compositions (\Cref{subsec:compatsecpasses}), and it gives additional technical insights on the secure compilation proofs (\Cref{subsec:seccompproofs}).

@@

The previous theorems talk about preserving properties expressed in the trace model of the languages of each compiler.
However, these trace models are not the same trace model we used to specify the properties of \Cref{sec:compprop} (which we denote with \ev{L}), and which serves as the ``ground truth'' of the meaning of a certain property.
% 
To bridge this gap, the formal development requires specifying additional trace relations, from each of the language trace models to the \ev{L} trace model, that, for example, relate $\src{Get\ \loc\ n}$ and $\src{Set\ \loc\ n}$ to $\ev{Use\ \loc\ n}$.
% 
One key insight of these relations is that they all omit context-made actions for two reasons: (1) contexts (which are universally quantified) can trivially invalidate any property and (2) we are interested in the component upholding the properties.
%

@@

\begin{theorem}[Properties Relation Correctness]\label{thm:prop-rel-corr}
  \begin{align*}
    \forall& \pi \in \{\tmssafe,\smssafe,\scctsafe,\sssafe\}, 
    \\
    \text{ for each }& \text{pair of languages } \src{L} \text{ and } \trg{L} \text{ used by the compilers},
    \\
    \text{ if }& \sigma_{\sim^{\trg{L}}_{\ev{L}}}(\pi) \sim^{\trg{L}}_{\ev{L}} \pi
    \text{ and } \sigma_{\sim^{\src{L}}_{\trg{L}}}(\trg{\pi}) \sim^{\src{L}}_{\trg{L}} \sigma_{\sim^{\trg{L}}_{\ev{L}}}(\pi)
    \\
    \text{ then }& \sigma_{\sim^{\src{L}}_{\trg{L}}}(\trg{\pi}) \sim^{\src{L}}_{\ev{L}} \pi
  \end{align*}
\end{theorem}
The complexity of this theorem is that the relation in the conclusion cannot be obtained by composing the two relations in the premises.
% 
What happens in practice is that programmers are given the relations from each language to \ev{L} and they need to build a compiler that induces a relation $\sim^{\src{L}}_{\trg{L}}$ that respects this theorem.
% 
% We now informally argue why this theorem holds in all of its specialised cases.

\paragraph{$\sigma_{\sim^{\Ltms}_{\Lms}}\left(\mssafe\cap\scctsafe\cap\sssafe\right)$}
Consider a trace $\irl{\trace}\in\mssafe$ and remember that $\sim^{\Ltms}_{\Lms}$ is the composition of $\sim^{\Ltms}_{\Ltrg}$ and $\sim^{\Ltrg}_{\Lms}$.
{\em All} traces $\trg{\trace}$ with $\trg{\trace}\sim^{\Ltrg}_{\Lms}\irl{\trace}$ are almost identical to $\irl{\trace}$. 
They differ in memory accesses and crashes, where $\irl{\lightning}$ can be related to a $\trg{\lightning}$ or a memory event, e.g., $\trg{Get\ \loc\ 2}$, for an out-of-bounds access.
In both cases, the translated event does not represent an out-of-bounds memory access.

Recall that $\sim^{\Ltms}_{\Ltrg}$ is an equality, so if $\src{\trace}\sim^{\Ltms}_{\Ltrg}\trg{\trace}$, we know that $\src{\trace}$ and $\trg{\trace}$ are identical.
% \footnote{Both $\sim^{\Ltms}_{\Ltrg}$ and $\sim^{\Ltrg}_{\Lms}$ filter context actions.}.

Now, starting from $\mssafe$ expressed in $\Lms$, we check that its translation in $\Ltms$ is still $\mssafe$.
% 
We can see that the property resulting from the projection of $\Lms$ traces back to $\Ltms$ faithfully models $\mssafe$, which translates to both $\scctsafe$ and $\sssafe$, 
\MPin{ equiv  +  other rel only adds traces that fail }
since these properties trivially hold for $\Ltms$, $\Ltrg$, and $\Lms$ traces.

\paragraph{$\sigma_{\sim^{\Lms}_{\Lscct}}\left(\mssafe\cap\scctsafe\cap\sssafe\right)$}

Consider a trace $\obj{\trace}\in\scctsafe$ and all $\irl{\trace}$ such that $\irl{\trace}\sim^{\Lms}_{\Lscct}\obj{\trace}$. 
In turn, $\obj{\trace}$ cannot contain any leaks of secret values, so it cannot violate $\scctsafe$.
Toggling data-independent timing mode has no impact on $\mssafe$ or $\sssafe$.

\paragraph{$\sigma_{\sim^{\Lscct}_{\Lspec}}\left(\mssafe\cap\scctsafe\cap\sssafe\right)$}

The argument here is analogous to the previous one, introducing $\ird{barrier}$ never violates $\sssafe$ and has no effect on $\mssafe$ or $\scctsafe$.

@@

\Cref{thm:rtpsim:sig} is always applicable, up to well-formedness of $\sim_1$ w.r.t. $\trg{\class[_2]}$.
However, consider $\ccb$ and $\ccspec$.
Even though swapping these compilers in the compilation chain is syntactically impossible, let us pretend that it is possible by just extending $\ccb$ for the constructs present in $\Lscct$ that are not present in $\Ltrg$.
Clearly, as evidenced earlier (\Cref{subsec:formalities:props}), $\ccb\cdot\ccspec$ is an acceptable composition that yields an interesting class of security properties, i.e., $\sigma_{\sim_{\irl{ms}}\bullet\sim_{\ird{\mathghost}}}\left(\mssafe\cap\scctsafe\cap\sssafe\right)$. 
However, swapping them $\ccspec\cdot\ccb$ means that $\ccb$ inserts new branches into the code that are not necessarily protected by a speculative execution barrier! 
It is crucial to ensure that the resulting class is meaningful, much similar to ensuring that a definition models what the formal methods expert intends to model. 
For the $\ccspec\cdot\ccb$ case, the class is $\sigma_{\sim_{\ird{\mathghost}}\bullet\sim_{\irl{ms}}}\left(\mssafe\cap\scctsafe\cap\sssafe\right)$
Since $\ev{Use\ \loc\ n}\sim_{\irl{ms}}\ev{Branch n}\cdot\ev{Spec}\cdots$, where $\cdots$ does not contain a $\ird{\ev{Barrier}}$ event, the resulting class of properties is only a subset of the intended one.
Because of this, it is crucial to analyze the precise shape of the class of properties after mapping a target-level property to a source-level property\footnote{Dito for mapping a source-level property to a target-level property.}.

@@

The composition theorems can only be used on compilers which have been proven secure in the sense that they robustly preserve a certain class of properties. 
Traditional proof techniques apply, the usual strategy is to establish a backwards simulation between source components and their compiled target counterpart. 
For the universally quantified target context, a technique known as backtranslation is well-established, which constructs a source-level context either by backtranslating the target-level context, amounting to a ``decompilation'', or by constructing it using the target-level trace, so that the source-context emits parts of the whole execution trace that are related to the parts emitted by the target-level context. 
Up to backtranslations for $\ccdce$ and $\cccf$, all backtranslations in the case-study are trace based, while for $\ccdce$ and $\cccf$, they are context-based. 
Since these optimisations have no significant semantic effect and source- and target-language are equal, the backtranslation amounts to an identity translation.

>>>>

This paper uses a general notion of robust preservation~\cite{abate2021extacc} that considers compilers that use languages with different trace models. 
To this end, considering a source trace $\src{\trace}$ and a target trace $\trg{\trace}$, there is a relation ($\sim$) describing the effect of a corresponding compiler (see \Cref{sec:casestud:rtp}). 
This relation induces the following two projection functions~\cite{abate2021extacc}: (1) the \emph{existential image} $\tau_\sim\left(\src{\pi}\right)$ and (2) the \emph{universal image} $\sigma_\sim\left(\trg{\pi}\right)$.
These projections map source-level (resp. target-level) properties to target-level (resp. source-level) properties in a way that identifies the ``same'' property across languages. 
The case study of this paper uses the universal image, since some considered properties, such as \gls*{ss}, are not definable in a higher-level language that, e.g., does not model speculation.
% For self-containment, the universal image is defined as follows~\cite{abate2021extacc}:

\begin{definition}[Universal Image]
% [Existential and Universal Image]
\label{def:universal:img}\label{def:sigma}
  % \[
  %   \tau_\sim\left(\src{\pi}\right) := 
  %     \left\{ 
  %       \trg{\trace} \mid \exists \src{\trace}\ldotp \src{\trace}\sim\trg{\trace}, \text{ and } \src{\trace}\in\src{\pi} 
  %     \right\}
  % \]
  \[ 
    \sigma_\sim\left(\trg{\pi}\right) := 
      \left\{ 
        \src{\trace} \mid \forall \trg{\trace}\ldotp \text{if }\src{\trace}\sim\trg{\trace}, \text{ then } \trg{\trace}\in\trg{\pi} 
      \right\}
  \]
\end{definition}

@@

\subsubsection{\gls*{ms}, Strict Cryptographic Constant Time}\label{sec:msscct-rel}

The combination of \gls*{ms} and \gls*{scct} is the intersection of these properties, \gls*{msscct}.
However, \gls*{ms} uses a different trace model than \gls*{scct}, so intersecting them would trivially yield the empty set. 
To remedy this issue, we introduce $\sim_{\ctsafe}: \preevent[_{\ctsafe}] \times \preevent[_{\mssafe}] $, a cross-language trace relation (whose key cases are presented below), that we use to intuitively unify the trace model in which the two properties are expressed:
\[
  \typerulenolabel{mscctrel:drop:tag}{}{\preevent[_{\mssafe}]\sim_{\ctsafe}\preevent[_{\mssafe}];\unlock{}}
  \typerulenolabel{mscctrel:drop:crash}{}{\lightning\sim_{\ctsafe}\lightning}
\]
\[
  \typerulenolabel{mscctrel:drop:branch}{}{\emptyevent\sim_{\ctsafe}\ev{Branch n};\securitytag{}}
  \typerulenolabel{mscctrel:drop:binop}{}{\emptyevent\sim_{\ctsafe}\ev{Binop n};\securitytag{}}
\]

Essentially, $\sim_{\ctsafe}$ ignores both the new $\ev{Branch\ n}$ and $\ev{Binop\ n}$ base-events as it relates security-insensitive actions ($\unlock$) to their equivalent counterparts.
Thus, traces of \gls*{ms} trivially satisfy \gls*{scct}.

@@

\subsubsection{Speculation Memory Safety}\label{sec:spec-ms-rel}

@@

{
\[
  \typerulenolabel{xrel:sms:read}{\trg{n}\text{ in bounds}}{\trg{Get\ \loc\ n;\comp}\sim^{\Ltrg}_{\Lms}\irl{Get\ \loc\ n;\comp}}
\]
% \[
%   \typerulenolabel{xrel:sms:panic}{\trg{n}\text{ out of bounds}}{\trg{Get\ \loc\ n;\comp}\sim^{\Ltrg}_{\Lms}\irl{\lightning}}
% \]
}

@@

This section first discusses how to connect each language-specific security property to the general security properties presented in \Cref{sec:compprop} (\Cref{subsec:formalities:maps}), then it demonstrates that the security property resulting from the universal image projection is faithful to the original property (\Cref{subsec:formalities:props}). 
Finally, this section discusses why the order of compiler passes matters, and how does our framework help with identifying insecure compositions (\Cref{subsec:compatsecpasses}), and it gives additional technical insights on the secure compilation proofs (\Cref{subsec:seccompproofs}).

@@

The previous theorems talk about preserving properties expressed in the trace model of the languages of each compiler.
However, these trace models are not the same trace model we used to specify the properties of \Cref{sec:compprop} (indicated with \ev{L}), which serves as the ``ground truth'' for the meaning of our properties.
% 
To bridge this gap, the formal development requires specifying additional trace relations, from each of the language trace models to the \ev{L} one, that, for example, relate $\src{Get\ \loc\ n}$ and $\src{Set\ \loc\ n}$ to $\ev{Use\ \loc\ n}$ (and that induce a related universal image that we use in \Cref{subsec:formalities:props}).
% 
One key insight of these relations is that they all omit context-made actions for two reasons: (1) contexts (which are universally quantified) can trivially invalidate any property and (2) we are interested in the component upholding the properties.
%

@@

\begin{theorem}[Properties Relation Correctness]\label{thm:prop-rel-corr}
  \begin{align*}
    \forall& \pi \in \{\tmssafe,\smssafe,\scctsafe,\sssafe\}, 
    \\
    \text{ for each }& \text{pair of languages } \src{L} \text{ and } \trg{L} \text{ used by the compilers},
    \\
    \text{ if }& 
    % \tau_{\sim^{\trg{L}}_{\ev{L}}}(\pi) \sim^{\trg{L}}_{\ev{L}} \pi
    \sigma_{\sim^{\trg{L}}_{\ev{L}}}(\pi) \sim^{\trg{L}}_{\ev{L}} \pi
    \text{ and } 
    % \sigma_{\sim^{\src{L}}_{\trg{L}}}(\tau_{\sim^{\trg{L}}_{\ev{L}}}(\pi)) \sim^{\src{L}}_{\trg{L}} \tau_{\sim^{\trg{L}}_{\ev{L}}}(\pi)
    \sigma_{\sim^{\src{L}}_{\trg{L}}}(\sigma_{\sim^{\trg{L}}_{\ev{L}}}(\pi)) \sim^{\src{L}}_{\trg{L}} \sigma_{\sim^{\trg{L}}_{\ev{L}}}(\pi)
    \\
    \text{ then }& 
    % \tau_{\sim^{\src{L}}_{\ev{L}}}(\pi) = \sigma_{\sim^{\src{L}}_{\trg{L}}}(\tau_{\sim^{\trg{L}}_{\ev{L}}}(\pi)) 
    \sigma_{\sim^{\src{L}}_{\trg{L}}}(\sigma_{\sim^{\trg{L}}_{\ev{L}}}(\pi)) \sim^{\src{L}}_{\ev{L}} \pi
  \end{align*}
\end{theorem}
The complexity of this theorem is that the relation in the conclusion cannot be obtained by composing the two relations in the premises.
% 
What happens in practice is that programmers are given the relations from each language to \ev{L} and they need to build a compiler that induces a relation $\sim^{\src{L}}_{\trg{L}}$ that respects this theorem.
% 
% We now informally argue why this theorem holds in all of its specialised cases.

\paragraph{$\sigma_{\sim^{\Ltms}_{\Lms}}\left(\tmssafe\cap\smssafe\cap\scctsafe\cap\sssafe\right)$}

For the four properties considered here, the trace models of $\Ltms$, $\Ltrg$, and $\Lms$ do not consider actions related to $\scctsafe$ and $\sssafe$, so these two properties are trivially translated correctly.
% 
We now discuss the remaining $\tmssafe$ and $\smssafe$ in the form of their intersection $\mssafe$.

Recall that $\sim^{\Ltms}_{\Lms}$ is the composition of $\sim^{\Ltms}_{\Ltrg}$ and $\sim^{\Ltms}_{\Lms}$.
% 
Since $\sim^{\Ltms}_{\Ltrg}$ is an equality, this relation trivially preserves the meaning of translated properties: related traces are identical!

Finally, let us consider a trace $\irl{\trace}\in\mssafe$ and understand what is that is related to via $\sim^{\Ltrg}_{\Lms}$.
% 
{\em All} traces $\trg{\trace}$ with $\trg{\trace}\sim^{\Ltrg}_{\Lms}\irl{\trace}$ are identical to $\irl{\trace}$ except for get and set actions, which require for in-bound accesses: this clearly respect $\mssafe$.
%

% In both cases, the translated event does not represent an out-of-bounds memory access.

% \footnote{Both $\sim^{\Ltms}_{\Ltrg}$ and $\sim^{\Ltrg}_{\Lms}$ filter context actions.}.

% Now, starting from $\mssafe$ expressed in $\Lms$, we check that its translation in $\Ltms$ is still $\mssafe$.
% % 
% We can see that the property resulting from the projection of $\Lms$ traces back to $\Ltms$ faithfully models $\mssafe$, which translates to both $\scctsafe$ and $\sssafe$, 
% \MPin{ equiv  +  other rel only adds traces that fail }
% since these properties trivially hold for $\Ltms$, $\Ltrg$, and $\Lms$ traces.

\paragraph{$\sigma_{\sim^{\Lms}_{\Lscct}}\left(\tmssafe\cap\smssafe\cap\scctsafe\cap\sssafe\right)$}

As before, the trace models of $\Lscct$ and $\Lms$ cannot express $\sssafe$, so that is trivially translated correctly.
% 
Also, the trace model of $\Lscct$ extends the one of $\Lms$ with respect to $\tmssafe$ and $\smssafe$ events, so the translation argument regarding those two properties is the same as before.
% 
Thus, we need to reason about whether $\scctsafe$ is translated correctly.

By definition, $\sim^{\Lms}_{\Lscct}$ only relates $\obj{\unlock}$ events, which is also the same relation induced by $\sim_{\ctsafe}$ in \Cref{sec:msscct-rel}.
% 
This ensures that composing the relations only relates $\unlock$ events, and thus the property is translated correctly.

% Consider a trace $\obj{\trace}\in\scctsafe$ and all $\irl{\trace}$ such that $\irl{\trace}\sim^{\Lms}_{\Lscct}\obj{\trace}$. 
% In turn, $\obj{\trace}$ cannot contain any leaks of secret values, so it cannot violate $\scctsafe$.
% Toggling data-independent timing mode has no impact on $\mssafe$ or $\sssafe$.

\paragraph{$\sigma_{\sim^{\Lscct}_{\Lspec}}\left(\tmssafe\cap\smssafe\cap\scctsafe\cap\sssafe\right)$}
The trace model of $\Lspec$ extends the one of $\Lscct$ with respect to $\tmssafe$, $\smssafe$, and $\scctsafe$, so the translation argument for those three properties is the same as before.
% 
Concerning $\sssafe$, $\sim^{\Lscct}_{\Lspec}$ relates $\Lspec$ speculation traces with $\Lscct$ branches with the $\lock_{\text{NONE}}$ tag, so the property is translated correctly, according to the relation defined in \Cref{sec:spec-ms-rel}.
 % introducing $\ird{barrier}$ never violates $\sssafe$ and has no effect on $\mssafe$ or $\scctsafe$.

@@

% \Cref{thm:rtpsim:sig} is applicable, up to the well-formedness of $\sim_1$ w.r.t. $\trg{\class[_2]}$.
Consider applying $\ccspec$ first and then $\ccb$ (albeit currently syntactically impossible).
% , let us pretend that it is possible by just extending $\ccb$ for the constructs present in $\Lscct$ that are not present in $\Ltrg$.
% Clearly, as evidenced earlier (\Cref{subsec:formalities:props}), $\ccb\cdot\ccspec$ is an acceptable composition that yields an interesting class of security properties, i.e., $\sigma_{\sim_{\irl{ms}}\bullet\sim_{\ird{\mathghost}}}\left(\mssafe\cap\scctsafe\cap\sssafe\right)$. 
In this case, $\ccb$ would insert new branches into the code that are not protected by a speculation barrier! 
This concern is reflected in the proofs that establish that the security class resulting of the composition of the trace relations is meaningful.
% , much similar to ensuring that a definition models what the formal methods expert intends to model. 
For the $\ccspec\cdot\ccb$ case, the class is $\sigma_{\sim_{\ird{\mathghost}}\bullet\sim_{\irl{ms}}}\left(\mssafe\cap\scctsafe\cap\sssafe\right)$.
Since $\ev{Use\ \loc\ n}\sim_{\irl{ms}}\irl{Branch\ n}\cdot\irl{Spec}\cdots$, where $\cdots$ does not contain a $\ird{\ev{Barrier}}$ event, the resulting class is \emph{not} the original $\sssafe$ that is intended and it would break the corresponding \Cref{thm:prop-rel-corr}.
% Because of this, it is crucial to analyze the precise shape of the class of properties after mapping a target-level property to a source-level property\footnote{Dito for mapping a source-level property to a target-level property.}.

@@

Our secure compilation proofs rely on established proof techniques called backtranslation~\cite{abate2019jour,patrignani2021rsc}, which let one construct a source context starting from either target traces (aka trace-based backtranslations) or target contexts (aka context-based backtranslations).
These backtranslations also require setting up cross-language relations between various language elements such as expressions and program states, so we leave these details for the technical reports.
% , the usual strategy is to establish a backwards simulation between source components and their compiled target counterpart. 
% For the universally quantified target context, a technique known as backtranslation is well-established, which constructs a source-level context either by backtranslating the target-level context, amounting to a ``decompilation'', or by constructing it using the target-level trace, so that the source-context emits parts of the whole execution trace that are related to the parts emitted by the target-level context. 
All backtranslations in the case-study are trace based except for those required by $\ccdce$ and $\cccf$, which are context-based (and they are essentially an identity function). 
% Since these optimisations have no significant semantic effect and source- and target-language are equal, the backtranslation amounts to an identity translation.
